Leadership across cultures and contexts: an exploratory analysis

### 2020-05-15 leadership_across_cultures_contexts.Rmd
Adding csl files.
Revisions in intro and functions intro sections.

---
title: "Leadership across cultures and contexts: an exploratory analysis"
author: "Zachary H. Garfield, Kristen L. Syme, and Edward H. Hagen"
date: "4/25/2019"
output: 
  bookdown::html_document2:
    number_sections: false
bibliography: Leader2.bib
csl: evolution-and-human-behavior.csl
---

@@

Processes of leadership and followership, which are central to social life in all societies, nevertheless vary from the informal roles observed in nonindustrial societies and smaller-scale organizations to the formal roles observed among WEIRD societies and their large-scale, complex organizations. Many researchers have turned to evolutionary theory to better understand this diversity. Evolutionary theories of leadership, in turn, draw on ethnographic cases of more traditional societies thought to more closely resemble the smaller-scale, face-to-face societies in which humans evolved. Currently, though, there is limited systematic data on the nature of leadership in such societies.

We build on our previous cross-cultural findings to systematically evaluate dimensions of leadership across a broad range of human cultural diversity. Over 1200 ethnographic texts from 59 mostly non-WEIRD populations in Human Relations Area Files (HRAF) were coded on leader qualities and functions. Because adaptations for leadership and followership could only have evolved if the fitness benefits exceeded the costs, we coded the costs and benefits for both leaders and followers across diverse contexts. We discovered evidence for both cultural universals in leadership, as well as important variation. Leaders in the ethnographic record were most commonly described as intelligent and knowledgeable, of high social status, accomplished, wealthy, and prosocial. The primary benefits leaders offered to followers were resolving conflicts, organizing cooperation, providing social functions, leading in inter-group conflicts, providing ritual services, and distributing resources. The primary benefits leaders received from followers included material resources, increased social status, and social services. The most frequent costs of leadership included loss of social status, resources or social services and increased risk of harm or conflict. For followers, the most frequent costs included loss of resources, costs in providing social services, and increased risk of harm or conflict.

@@

Leaders substantially shape group structures and outcomes across cultures. The structure of groups and their goals, in turn, as well as their subsistence strategies, can favor particular leader qualities. Despite decades of research and theorizing on leadership in numerous social science disciplines, however, to our knowledge there is no comprehensive or systematic investigation of leader functions and qualities, and the costs and benefits of leadership and followership, across a representative sample of nonindustrial societies. Moreover, outside of anthropology, most research and theorizing on leadership is based on data from Western or Westernized, postindustrial societies [@von_rueden_leadership_2015].

@@

There is now well established consensus within the behavioral sciences on the importance of data from the full range of human cultural diversity. Systemic biases in much of social science are well known, and might be due, in part, to the lack of data from populations other than American undergraduates [@henrich_weirdest_2010, *The weirdest people in the world?*]. Since the seminal WEIRD people paper, many social scientists have taken up the challenge to diversify their samples and have sought out "non-WEIRD" participants and databases. Dichotomizing populations and their psychologies and social dynamics as WEIRD or non-WEIRD, however, is a mistake.^[A Google Scholar search for "non-WEIRD" produced over 1000 results.] As anthropology has conclusively demonstrated, there is enormous diversity across societies that does not remotely resemble a dichotomy.

@@

<!-- EVOLUTIONARY THEORY APPROACH -->
Substantial scientific work across psychology, sociology, political science, and managerial studies has focused on trying to uncover interrelationships of dimensions of leadership, including the traits of leaders, their roles, and relationships with culture, economics, and group structure. Until recently, this large body of work was relatively disconnected from approaches in anthropology and biology. Under a unifying evolutionary framework, social and biological scientists have now produced a sizable body of theory on the evolution and cross-cultural patterning of leadership and followership [e.g., @price_evolution_2014; @henrich_big_2015; @richerson_tribal_2009; @van_vugt_evolutionary_2006; @pietraszewski_evolution_2019; @hooper_theory_2010; @kaplan_evolutionary_2009; @von_rueden_roots_2014; @hagen_leadership_2019], accompanied by a growing body of empirical results [e.g., @bowser_womens_2010; @alami_mothers_2020; @cheng_two_2013; @wiessner_collective_2019; @redhead_dynamics_2018; @garfield_investigating_2019; @garfield_evolutionary_2019; @von_rueden_leadership_2014; @von_rueden_sex_2018; @smith_leadership_2016; @smith_obstacles_2018; @lonati_what_2019]. This integration as been informative [see @van_vugt_genes_2020].

<!-- GROUP CONTEXT and cost PROBLEMS -->
Much of leadership studies, including evolutionary approaches, have focused on community or political leaders, often overlooking subtleties in the nature of leadership and followership across social contexts and levels of social organization, such as among kin groups, economic groups, and sub-groupings of individuals within residential communities. Lastly, while the benefits leaders accrue for their services have received substantial attention -- fitness benefits in particular -- the costs of leadership for both leaders and followers have been relatively under-investigated [for discussions on costs of leadership see @cheng_dominance_2019; @price_evolution_2014; @glowacki_leadership_2015; @von_rueden_leadership_2014]. Outstanding questions for evolutionary leadership scholars include the degree to which the nature of leadership is associated with between-group population-level variation and within-group variation in social contexts, as well as identifying the associated benefits *and* costs both leaders and followers incur.
<!-- CAN PULL SOME COST CITATIONS FROM NHB PAPER -->

@@

<!-- CLOSING -->
We first briefly review theoretical, ethnographic, and empirical perspectives on the qualities of leaders, the functions they serve, and the costs and benefits of leadership across human societies, including WEIRD and non-WEIRD populations. We then outline the aims and methods of the current cross-cultural study -- which focuses on uncovering patterns among non-WEIRD populations -- before presenting results and interpretations. Through novel exploratory analyses encompassing the full range of cultural diversity and diverse social contexts we aim to advance leadership studies across disciplines and provide researchers with a foundation for more detailed approaches to human leadership across cultures and contexts.

@@

Theoretical literature suggests these demands most commonly arise in the contexts of collective actions and in conflict resolution [e.g., @gavrilets_convergence_2016; @hooper_theory_2010; @von_rueden_solving_2015; @henrich_big_2015; @pietraszewski_evolution_2019]. Demands of mediation in within-group conflict resolution, although variable in scale, are probably a universal feature of human groups [@king_origins_2009; @van_vugt_leadership_2008]. In the absence of institutionalized organizational structure leaders often take on essential roles in within-group coordination and collective action, including facilitating conflict resolution within or between autonomous households or kin groups, managing individual investments to group-wide projects, and administering sanctions to individuals who fail to appropriately contribute to collective actions, comply with group policies, or seriously violate social norms. Demands of between-group interactions, including both conflict and cooperation are also very common across small-scale societies, including hunter-gatherers [@glowacki_evolutionary_2017; @hames_pacifying_2019; @richerson_cultural_2016; @henrich_evolution_2009; @apicella_social_2012]. Across relatively small-scale societies representing the group and managing both internal and external relationships and conflicts are expected to be the most salient functions of community leaders. Within communities, however, leaders may emerge whenever individuals congregate. The role of leaders in ritual contexts, such as in initiations [@singh_cultural_2017], and in providing social services, including facilitating marriages [@walker_evolutionary_2011] and organizing social activities [@wiessner_food_1996], are also expected to be common functions of leaders.

@@

<!-- WEIRD ORGNIZATIONAL STRUCTURE TELLS FUNCTIONS -->
In WEIRD societies, community and political leaders are primarily heads of governmental and corporate bodies [@bass_handbook_1990]. Other public service, nonprofit leaders and managers include law enforcement and police officers, college presidents and school principles, and grassroots community leaders, to name a few [@martiskainen_role_2017; @azzam_community_2003; @grace_school_2005]. These positions typically are associated with institutionalized responsibilities and specific functions. The structure of such organizations are generally hierarchical with nested bodies and an internal 'chain of command.' At higher levels of large-scale organizations, leaders primarily serve as representatives in inter-group interactions, facilitate the creation and adoption of policies, and serve as figure heads embodying the strategic vision of the organization [@keohane_thinking_2010]. They are not typically directly involved in resolving interpersonal conflicts between individuals, however, they are often the final end-point in a decision making hierarchy and the results of a single decision of these leaders can trickle down through industry or government with global ramifications.

<!-- THEORY ON WEIRD FUNCTIONS -->
Scholars in managerial studies have developed multiple models on the functions of leadership. Classic management theorists [e.g., @davis_learning_1951] emphasized the planning, organizing, and controlling functions of leaders in formal organizational and business settings [see @bass_handbook_1990]. These perspectives focused on the rational, mechanistic aspects of group functioning and often overlooked interpersonal dynamics. Later theorists incorporated cognitive and socioemotional components of leaders functions. For example, @selznick_leadership_1957 describes the functions of organization leaders is to define the goals of the organization, create structures for accomplishing group goals, defend institutional integrity, and maintain internal consistency among group members.

@@

<!-- COMPARATIVE DATA ON WEIRD FUNCTIONS -->
More recently, many managerial scholars have adopted integrative and comparative approaches to better understand cultural or cross-national diversity in leadership systems. @behrendt_integrative_2017 draw on empirical results from the managerial literature and suggest the behaviors of leaders across managerial contexts are classifiable as either *task-oriented*, including enhancing understanding, strengthening motivation and facilitating implementation, or *relation-oriented*, including fostering coordination, promoting cooperation and mobilizing resources. This framework is intended to provide a more fine-grained and externally valid view of leadership functions, compared to traditional models of transactional leadership, which tend to oversimplify dimensions of leader-follower dynamics. More integrated and holistic models have also allowed managerial scholars to develop comparative theoretical frameworks for leader functioning. Culturally Endorsed Implicit Leadership Theory (CLT), suggests cultural variation in social norms has the potential to impact follower preferences for leadership styles [@moore_literature_2018]. The Global Leadership and Organizational Behavior Effectiveness (GLOBE) project is a large-scale international research program aimed at understanding leader effectiveness and cultural variation across large organizations. The GLOBE project has produced multiple cross-national data sets on leadership and hundreds of publications. For example, *The GLOBE Study of CEO Leadership Behavior and Effectiveness in 24 Countries* draws on data from over 1,000 CEOs and over 5,000 senior executives from corporations in varied industries across 24 countries [@house2013strategic]. The goals of the GLOBE CEO Study were to address (1) how national culture influences expectations for the behaviors of leaders, (2) which behaviors of CEOs are typically associated with organizational success, (3) differences between high-functioning and low-functioning CEOs, and (4) the importance of CEO behaviors aligning with societal expectations. Through path analysis the study concluded that cultural values indirectly predict CEO leadership behavior through culturally endorsed leadership expectations. That is, @house2013strategic suggest leaders moderate their behavior through culturally learned values and leadership-specific values.

The GLOBE CEO Study defined six "global leadership behaviors" developed from 21 "primary leadership behaviors." Among the global leadership behaviors, they found that charismatic leadership behavior was consistently the most important behavioral trait regarding firm performance and dedication of top management team (TMT) members, followed by team-oriented behavior and humane-oriented behavior. Participative leadership was moderately related to TMT dedication, however was not related to firm performance. The most successful CEOs are those who are rated as more visionary, performance oriented, and decisive. Furthermore, results strongly suggest the fit between a CEO's behavioral profile and the expectations of leadership within a society strongly predict TMT dedication and firm performance. By developing higher-order features of leadership behavior and situating leadership functions within more detailed classification frameworks, insights from managerial studies can more effectively be coupled with approaches from across behavioral science.

<!-- ADDED CONCLUSION PARAGRAPH/STATEMENT.  -->
In contrasting the functions leaders serve in smaller-scale, non-WEIRD populations from larger-scale, WEIRD ones, community leaders in non-WEIRD settings are more likely to serve context-specific functions based on role-specialization [see @smith_leadership_2016]. The hunt leader does not necessarily have authority over military raids and the lineage head does not necessarily have authority over ritual ceremonies. In WEIRD settings however, community, organizational, and political leaders often have decision-making authority across multiple contexts. In the United States, despite a designed separation of power across branches of government, presidents, governors, and mayors oversee governmental bodies and have substantial influence in economics, social and legal policies, and international and inter-regional relations. Community leadership functions in WEIRDer populations will likely be increasingly focused on economic contexts, creating and managing organizational structures, and inter-group relations, whereas in less WEIRD contexts community leaders will primarily function to resolve interpersonal conflicts and manage intra-group social dynamics.
<!-- Could add ethnographic examples for SSS and LSS  -->

>>>>

---
title: "Leadership across cultures and contexts: an exploratory analysis"
author: "Zachary H. Garfield, Kristen L. Syme, and Edward H. Hagen"
date: "4/25/2019"
output: 
  bookdown::html_document2:
    number_sections: false
bibliography: Leader2.bib
csl: apa-6th-edition.csl
---

@@

Processes of leadership and followership, which are central to social life in all societies, nevertheless vary from the informal roles observed in nonindustrial societies and smaller-scale organizations to the formal roles observed among WEIRD societies and their complex organizations. Many researchers have turned to evolutionary theory to better understand this diversity. Evolutionary theories of leadership, in turn, draw on ethnographic cases of societies thought to more closely resemble the smaller-scale, face-to-face communities in which humans evolved. Currently, though, there is limited systematic data on the nature of leadership in such societies.

Building on our previous cross-cultural findings we systematically evaluate dimensions of leadership across a broad range of human cultural diversity. Over 1200 ethnographic texts from 59 mostly non-WEIRD populations in Human Relations Area Files (HRAF) were coded on leader qualities and functions. Because adaptations for leadership and followership could only have evolved if the fitness benefits exceeded the costs, we coded the costs and benefits for both leaders and followers across diverse contexts. We discovered evidence for both cultural universals in leadership, as well as important variation. Leaders in the ethnographic record were most commonly described as intelligent and knowledgeable, of high social status, accomplished, wealthy, and prosocial. The primary benefits leaders offered to followers were resolving conflicts, organizing cooperation, providing social functions, leading in inter-group conflicts, providing ritual services, and distributing resources. The primary benefits leaders received from followers included material resources, increased social status, and social services. The most frequent costs of leadership included loss of social status, resources or social services and increased risk of harm or conflict. For followers, the most frequent costs included loss of resources, costs in providing social services, and increased risk of harm or conflict.

@@

Leaders substantially shape group structures and outcomes across cultures. The structure of groups and their goals, in turn, as well as their subsistence strategies, can favor particular leader qualities. Despite decades of research and theorizing on leadership in numerous social science disciplines, to our knowledge there is no comprehensive or systematic investigation of leader functions and qualities, and the costs and benefits of leadership and followership, across a representative sample of nonindustrial societies. Moreover, outside of anthropology, most research and theorizing on leadership is based on data from Western or Westernized, postindustrial societies [@von_rueden_leadership_2015].

@@

There is now broad consensus within the behavioral sciences on the importance of data from the full range of human cultural diversity. Systemic biases in much of social science are well known, and might be due, in part, to the lack of data from populations other than American undergraduates [@henrich_weirdest_2010, *The weirdest people in the world?*]. Since the seminal WEIRD people paper, many social scientists have taken up the challenge to diversify their samples and have sought out "non-WEIRD" participants and databases. Dichotomizing populations and their psychologies and social dynamics as WEIRD or non-WEIRD, however, is a mistake.^[A Google Scholar search for "non-WEIRD" produced over 1000 results.] As anthropology has conclusively demonstrated, there is enormous diversity across societies that does not remotely resemble a dichotomy.

@@

<!-- EVOLUTIONARY THEORY APPROACH -->
Substantial scientific work across psychology, sociology, political science, and managerial studies has focused on trying to uncover interrelationships of dimensions of leadership, including the traits of leaders, their roles, and relationships with culture, economics, and group contex. Until recently, this large body of work was relatively disconnected from approaches in anthropology and biology. Under a unifying evolutionary framework, social and biological scientists have now produced a sizable body of theory on the evolution and cross-cultural patterning of leadership and followership [e.g., @price_evolution_2014; @henrich_big_2015; @richerson_tribal_2009; @van_vugt_evolutionary_2006; @pietraszewski_evolution_2019; @hooper_theory_2010; @kaplan_evolutionary_2009; @von_rueden_roots_2014; @hagen_leadership_2019], accompanied by a growing body of empirical results [e.g., @bowser_womens_2010; @alami_mothers_2020; @cheng_two_2013; @wiessner_collective_2019; @redhead_dynamics_2018; @garfield_investigating_2019; @garfield_evolutionary_2019; @von_rueden_leadership_2014; @von_rueden_sex_2018; @smith_leadership_2016; @smith_obstacles_2018; @lonati_what_2019]. This integration as been informative [see @van_vugt_genes_2020].

<!-- GROUP CONTEXT and cost PROBLEMS -->
Much of leadership studies, including evolutionary approaches, have focused on community or political leaders, often overlooking subtleties in the nature of leadership and followership across social contexts and levels of social organization, such as among kin groups, economic groups, and sub-groupings of individuals within residential communities. Lastly, while the benefits leaders accrue for their services have received substantial attention -- fitness benefits in particular -- the costs of leadership for both leaders and followers have been relatively under-investigated [for discussions on costs of leadership see @cheng_dominance_2019; @price_evolution_2014; @glowacki_leadership_2015; @von_rueden_leadership_2014]. Outstanding questions for evolutionary leadership scholars include the degree to which dimensions of leadership are associated with between-group population-level variation and within-group variation in social contexts, as well as identifying the associated benefits *and* costs both leaders and followers incur.
<!-- CAN PULL SOME COST CITATIONS FROM NHB PAPER -->

@@

<!-- CLOSING -->
We first briefly review theoretical, ethnographic, and empirical perspectives on the qualities of leaders, the functions they serve, and the costs and benefits of leadership across human societies, including WEIRD and non-WEIRD populations. We then outline the aims and methods of the current cross-cultural study -- which focuses on exploring patterns among non-WEIRD populations -- before presenting results and interpretations. Through novel exploratory analyses encompassing a broad range of cultural diversity and social contexts we aim to advance leadership studies across disciplines and provide researchers with a foundation for more detailed approaches to human leadership across cultures and contexts.

@@

Theoretical literature suggests these demands most commonly arise in the contexts of collective actions and in conflict resolution [e.g., @gavrilets_convergence_2016; @hooper_theory_2010; @von_rueden_solving_2015; @henrich_big_2015; @pietraszewski_evolution_2019]. Demands of mediation in within-group conflict resolution, although variable in scale, are probably a universal feature of human groups [@king_origins_2009; @van_vugt_leadership_2008]. In the absence of institutionalized organizational structure leaders often take on essential roles in within-group coordination and collective action, including facilitating conflict resolution within or between autonomous households or kin groups, managing individual investments to group projects, and administering sanctions to individuals who fail to appropriately contribute or seriously violate social norms. Demands of between-group interactions, including both conflict and cooperation are also very common across small-scale societies, including hunter-gatherers [@glowacki_evolutionary_2017; @hames_pacifying_2019; @richerson_cultural_2016; @henrich_evolution_2009; @apicella_social_2012]. Across relatively small-scale societies, representing the group and managing internal and external relationships and conflicts are expected to be the most salient functions of community leaders. Within communities, however, leaders may emerge whenever individuals congregate. The role of leaders in ritual contexts, such as in initiations [@singh_cultural_2017], and in providing social services including facilitating marriages [@walker_evolutionary_2011] and organizing social activities [@wiessner_food_1996], are also expected to be common functions of leaders.

@@

<!-- WEIRD ORGNIZATIONAL STRUCTURE TELLS FUNCTIONS -->
In WEIRD societies, community and political leaders are primarily heads of governmental and corporate bodies [@bass_handbook_1990]. Other public service, nonprofit leaders and managers include law enforcement and police officers, college presidents and school principles, and grassroots community leaders, to name a few [@martiskainen_role_2017; @azzam_community_2003; @grace_school_2005]. These positions typically are associated with institutionalized responsibilities and specific functions. The structure of such organizations are generally hierarchical with nested bodies and an internal 'chain of command.' At higher levels of large-scale organizations, leaders primarily serve as representatives in inter-group interactions, facilitate the creation and adoption of policies, and serve as figure heads embodying the strategic vision of the organization [@keohane_thinking_2010]. They are not typically directly involved in resolving interpersonal conflicts between individuals, however, they are often the final end-point in a decision making hierarchy and the results of a single decision of these leaders can trickle down through industry or government with global ramifications. Within local communities and in smaller-scale economic groups, however, leaders in WEIRD populations *are* likely to be direlty involved in mediating conflicts and managing local cooperative projects [e.g., @acheson1988lobster; @rich_dynamics_1980].

<!-- THEORY ON WEIRD FUNCTIONS -->
Scholars in managerial studies have developed multiple models on the functions of leadership. Classic management theorists [e.g., @davis_learning_1951] emphasized the planning, organizing, and controlling functions of leaders in formal organizational and business settings [see @bass_handbook_1990]. These perspectives focused on the rational, mechanistic aspects of group functioning and often overlooked interpersonal dynamics. Later theorists incorporated cognitive and socioemotional components of leaders functions. For example, @selznick_leadership_1957 describes the functions of organization leaders are to define the goals of the organization, create structures for accomplishing group goals, defend institutional integrity, and maintain internal consistency among group members.

@@

<!-- COMPARATIVE DATA ON WEIRD FUNCTIONS -->
More recently, many managerial scholars have adopted comparative approaches to better understand cultural or cross-national diversity in leadership systems. @behrendt_integrative_2017 draw on empirical results from the managerial literature and suggest the behaviors of leaders across managerial contexts are classifiable as either *task-oriented*, including enhancing understanding, strengthening motivation and facilitating implementation, or *relation-oriented*, including fostering coordination, promoting cooperation and mobilizing resources. This framework is intended to provide a more fine-grained and externally valid view of leadership functions, compared to traditional models of transactional leadership, which tend to oversimplify dimensions of leader-follower dynamics. More integrated models have also allowed managerial scholars to develop comparative theoretical frameworks for leader functioning. Culturally Endorsed Implicit Leadership Theory (CLT), suggests cultural variation in social norms has the potential to impact follower preferences for leadership styles [@moore_literature_2018]. The Global Leadership and Organizational Behavior Effectiveness (GLOBE) project is a large-scale international research program aimed at understanding leader effectiveness and cultural variation across large organizations. The GLOBE project has produced multiple cross-national data sets on leadership and hundreds of publications. For example, *The GLOBE Study of CEO Leadership Behavior and Effectiveness in 24 Countries* draws on data from over 1,000 CEOs and over 5,000 senior executives from corporations in varied industries across 24 countries [@house2013strategic]. The goals of the GLOBE CEO Study were to address (1) how national culture influences expectations for the behaviors of leaders, (2) which behaviors of CEOs are typically associated with organizational success, (3) differences between high-functioning and low-functioning CEOs, and (4) the importance of CEO behaviors aligning with societal expectations. Through path analysis the study concluded cultural values indirectly predict CEO leadership behavior through culturally endorsed leadership expectations. That is, @house2013strategic suggest leaders moderate their behavior through culturally learned values and leadership-specific values.

The GLOBE CEO Study defined six "global leadership behaviors" developed from 21 "primary leadership behaviors." Among the global leadership behaviors, they found charismatic leadership behavior was consistently the most important behavioral trait regarding firm performance and dedication of top management team (TMT) members, followed by team-oriented behavior and humane-oriented behavior. 
<!-- Participative leadership was moderately related to TMT dedication, however was not related to firm performance.  -->
The most successful CEOs are those who are rated as more visionary, performance oriented, and decisive. Furthermore, results strongly suggest the fit between a CEO's behavioral profile and the expectations of leadership within a society strongly predict TMT dedication and firm performance. By developing higher-order features of leadership behavior and situating leadership functions within more detailed frameworks, insights from managerial studies can more effectively be coupled with approaches from across behavioral science.

<!-- ADDED CONCLUSION PARAGRAPH/STATEMENT.  -->
In contrasting the functions leaders serve in smaller-scale, non-WEIRD populations from larger-scale, WEIRD ones, community leaders in non-WEIRD settings are more likely to serve context-specific functions based on role-specialization [see @smith_leadership_2016]. The hunt leader does not necessarily have authority over military raids and the lineage head does not necessarily have authority over ritual ceremonies. In WEIRD settings however, community, organizational, and political leaders often have decision-making authority across multiple contexts. 
<!-- In the United States, despite a designed separation of power across branches of government, presidents, governors, and mayors oversee governmental bodies and have substantial influence in economics, social and legal policies, and international and inter-regional relations.  -->
Community leadership functions in WEIRD populations will likely be increasingly focused on economic contexts, creating and managing organizational structures, and inter-group relations, whereas in less WEIRD contexts community leaders will primarily function to resolve interpersonal conflicts and manage intra-group social dynamics.
